\documentclass[11pt]{article}

\usepackage[utf8x]{inputenc}
\usepackage[L7x]{fontenc}  
\usepackage[pdftex]{graphicx}
\usepackage{fourier}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{wrapfig}
\usepackage{sectsty}
\usepackage{asymptote}
\usepackage{colonequals}
\usepackage{color}
\usepackage{calc}
\usepackage{etoolbox}

% COMMENT OUT THE SECOND LINE TO MAKE A HANDOUT WITHOUT SOLUTIONS
\newtoggle{solutions}
\toggletrue{solutions}

\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{theorem}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{exercise}{Exercise}

\sectionfont{\large}
\sectionfont{\normalsize}

\def\Arg{\mathop{\rm Arg}\nolimits}
\def\Res{\mathop{\rm Res}}
\renewcommand\Im{\mathop{\rm Im}\nolimits}
\renewcommand\Re{\mathop{\rm Re}\nolimits}
\newcommand\Arctan{\mathop{\rm Arctan}\nolimits}
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Z}{\mathbb{Z}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\Chat}{\hat{\mathbb{C}}}
\newcommand{\UHP}{\mathbb{H}}
\DeclareMathOperator{\area}{area}
\DeclareMathOperator{\dist}{dist}
\DeclareMathOperator{\interior}{int}
\DeclareMathOperator{\id}{id}

\pagestyle{empty}

\textwidth = 6.5 in
\textheight = 9 in
\oddsidemargin = -0.125 in
\evensidemargin = 0.0 in
\topmargin = -0.2 in
\headheight = 0.0 in
\headsep = 0.2 in
\parskip = 0.2 in
\parindent = 0.0 in

\def\inv{^{-1}}

\newcounter{prob}
	\setcounter{prob}{1}

\newcommand\itm{\theprob.  \stepcounter{prob}}

\iftoggle{solutions}{
\newcommand\sol[2]{\textit{Solution}. #1}
}
{
\newcommand\sol[2]{#2}
}


\newcommand{\problem}[1]{
\makebox[0.2cm]{\textbf{\itm}}  \begin{minipage}[t]{\linewidth-0.75cm}
#1
\end{minipage}
}

\begin{document}
\thispagestyle{empty}

\begin{center}
  18.022 Recitation Handout \iftoggle{solutions}{(with solutions)}{} \\
  10 September 2014 
\end{center}

\itm For each of the following pairs of vectors $\mathbf{a}$ and $\mathbf{b}$, calculate $ \mathbf{a}\cdot \mathbf{b} $ and $\|\mathbf{a}\|\|\mathbf{b}\|$. 

(a) $\mathbf{a}=(1,5)$ and $\mathbf{b}=(-2,3)$ 

\sol{$\mathbf{a}\cdot\mathbf{b} = 1\cdot-2 + 5\cdot 3 = 13$, and $\|\mathbf{a}\|\|\mathbf{b}\|=\sqrt{1^2+5^2}\sqrt{(-2)^2+3^2}=\sqrt{26\cdot 13}=13\sqrt{2}$}{}

(b) $\mathbf{a}=(3,-5)$ and $\mathbf{b}=(2,0)$ 

\sol{$\mathbf{a}\cdot\mathbf{b} = 3\cdot2 + -5\cdot 0 = 6$, and $\|\mathbf{a}\|\|\mathbf{b}\|=2\sqrt{17}$}{}

(c) $\mathbf{a}=(-2,4,1)$ and $\mathbf{b}=(4,1,2)$ 

\sol{$\mathbf{a}\cdot\mathbf{b} = -2\cdot4 + 4\cdot 1+1\cdot2 = -2$, and $\|\mathbf{a}\|\|\mathbf{b}\|=21.$}{}

(d) Conjecture an inequality relating $|\mathbf{a}\cdot \mathbf{b}|$ and $\|\mathbf{a}\|\|\mathbf{b}\|$ for $\mathbf{a},\mathbf{b}\in \R^n$.

\sol{In all three cases above, we see that $|\mathbf{a}\cdot \mathbf{b}|\leq \|\mathbf{a}\|\|\mathbf{b}\|$.}{}

(e) (Fun/Challenge problem) To prove the inequality conjectured in (d) (called the \textit{Cauchy-Schwarz inequality}), expand the left-hand side of the inequality $\|\mathbf{a}+\lambda\mathbf{b}\|^2\geq 0$, where $\lambda$ is any real number. 

\sol{Expanding $\|\mathbf{a}+\lambda\mathbf{b}\|^2\geq 0$ gives $\|\mathbf{a}\|^2+2\lambda \mathbf{a}\cdot \mathbf{b} + \lambda^2 \|\mathbf{b}\|^2 \geq 0$. Since we want to get the strongest possible inequality, we choose $\lambda$ to minimize the quadratic expression on the left-hand side. Substituting $\lambda = -\mathbf{a}\cdot\mathbf{b}/\|\mathbf{b}\|^2$, the resulting inequality simplifies to $|\mathbf{a}\cdot \mathbf{b}|\leq \|\mathbf{a}\|\|\mathbf{b}\|$.}
{
\vfill
}

\itm (1.3.20 in \textit{Colley}) Suppose that a force $\mathbf{F}=(1,-2)$ is acting on an object moving parallel to the vector $(4,1)$. Decompose $\mathbf{F}$ into a sum of vectors $\mathbf{F}_1$ and $\mathbf{F}_2$, where $\mathbf{F}_1$ points along the direction of motion and $\mathbf{F}_2$ is perpendicular to the direction of motion.

\sol{
  We project $(1,-2)$ onto $(4,1)$ using the formula $\text{proj}_\mathbf{b}\mathbf{a}=\left(\frac{\mathbf{a}\cdot \mathbf{b}}{\mathbf{b}\cdot \mathbf{b}}\right)\mathbf{b}$. We obtain $F_1=\frac{2}{17}(4,1)=(\frac{8}{17},\frac{2}{17})$ and $F_2=F_-F_2=(1,-2)-\left(\frac{8}{17},\frac{2}{17}\right)=\left(\frac{9}{17},-\frac{36}{17}\right)$. 
}{
\vfill
}

\itm (1.3.17 in \textit{Colley}) Is it ever the case that the projection of $\mathbf{a}$ onto $\mathbf{b}$ and the projection of $\mathbf{b}$ onto $\mathbf{a}$ are the same vector? If so, under what conditions? 

\sol{Suppose $\text{proj}_\mathbf{b}\mathbf{a}=\text{proj}_\mathbf{a}\mathbf{b}$. Since a projection onto $\mathbf{a}$ is either 0 or parallel to $\mathbf{a}$, we see that either both projections are zero (which happens if and only if $\mathbf{a}$ and $\mathbf{b}$ are orthogonal), or they are parallel. Moreover, if they are parallel, then in order to be equal they have to have the same length and direction. So $\text{proj}_\mathbf{b}\mathbf{a}=\text{proj}_\mathbf{a}\mathbf{b}$ if and only if $\mathbf{a}\cdot \mathbf{b} = 0$ or $\mathbf{a}=\mathbf{b}$.} { \vfill }

\newpage 

\itm (1.3.25 in \textit{Colley}) Use vectors to show that the diagonals of a parallelogram have the same length if and only if the parallelogram is a rectangle. (Hint: let $\mathbf{a}$ and $\mathbf{b}$ be vectors along two sides of the parallelogram, and express vectors running along the diagonals in terms of $\mathbf{a}$ and $\mathbf{b}$.) 

\sol{
  The diagonal vectors are $\mathbf{a}+\mathbf{b}$ and $\mathbf{a}-\mathbf{b}$. The lengths of these vectors are equal if and only if $\|\mathbf{a}+\mathbf{b}\|^2=\|\mathbf{a}-\mathbf{b}\|^2$. The left-hand side simplifies to $(\mathbf{a}+\mathbf{b})\cdot (\mathbf{a}+\mathbf{b})=\mathbf{a}\cdot \mathbf{a} + 2 \mathbf{a}\cdot \mathbf{b}+\mathbf{b}\cdot \mathbf{b}$. Simplifying the right-hand side similarly and canceling terms, we are left with $\mathbf{a}\cdot \mathbf{b}=0$, which says that the two sides of the parallelogram are perpendicular. 
}{
\vfill
}

\itm (1.3.23 in \textit{Colley}) Let $A$, $B$, and $C$ denote the vertices of a triangle. Let $0<r<1$. If $P_1$ is the point on $\overline{AB}$ located $r$ times the distance from $A$ to $B$ and $P_2$ is the point on $\overline{AC}$ located $r$ times the distance from $A$ to $C$, use vectors to show that $\overline{P_1P_2}$ is parallel to $\overline{BC}$ and has $r$ times the length of $\overline{BC}$. 

\sol{
This is an application of the distributive property of scalar multiplication across vector addition. Let $\mathbf{u}$ be the vector from $A$ to $B$, and let $\mathbf{v}$ be the vector from $A$ to $C$. Then the vector from $B$ to $C$ is $\mathbf{u}-\mathbf{v}$, and the vector from $P_1$ to $P_2$ is $r \mathbf{u}-r \mathbf{v}$. Reverse distributing, we write this as $r(\mathbf{u}-\mathbf{v})$, which says that $\overline{P_1P_2}$ is parallel to $\overline{BC}$ and has $r$ times the length of $\overline{BC}$. 
}{
\vfill
}

\itm (1975 USAMO) Let $A$, $B$, $C$, and $D$ be four points in $\mathbb{R}^3$. Use vectors to show that 
\[
AB^2 +BC^2 + CD^2 + DA^2 \geq AC^2 + BD^2. 
\]
(This generalizes the fact that the sum of the squares of the sides of a quadrilateral is at least the sum of the squares of its diagonals.) Make a statement about when equality holds. 

\sol{
Let $\mathbf{u}$ be the vector from $A$ to $B$, let $\mathbf{v}$ be the vector from $A$ to $D$, and Let $\mathbf{w}$ be the vector from $B$ to $C$. Then we have been asked to prove 
\[
\mathbf{u}\cdot \mathbf{u} + \mathbf{v}\cdot \mathbf{v} + \mathbf{w}\cdot \mathbf{w} + (\mathbf{u}+\mathbf{w}-\mathbf{v})\cdot (\mathbf{u}+\mathbf{w}-\mathbf{v}) \geq (\mathbf{u}+\mathbf{w})\cdot (\mathbf{u}+\mathbf{w}) + (\mathbf{u}-\mathbf{v})\cdot (\mathbf{u}-\mathbf{v})
\]
Distributing and simplifying, we see that this inequality is equivalent to $\|\mathbf{w}-\mathbf{v}\|^2 \geq 0$. This holds with equality if and only if $\mathbf{w}=\mathbf{v}$, i.e., if and only if the four points form a parallelogram. 
}{
\vfill
}

% \itm Consider a triangle $T$ with vertices $A=(1,0)$, $B=(0,1)$, and $C$ on the line $y=-x$. Use a cross product to find the area of $T$, and show that it does not depend on the location of $C$. 

% \sol{
% Let $C=(t,-t)$, where $t$ is any real number. Then the area is 1/2 times the absolute value of the cross product $\overrightarrow{AB}\times\overrightarrow{AC}$. Thus 
% \[
% \text{area}(\bigtriangleup ABC) = \frac{1}{2} \left|
% \begin{array}{ccc}
% \mathbf{i} & \mathbf{j} & \mathbf{k} \\
% 1 & -1 & 0 \\
% t & -t-1 & 0
% \end{array}
% \right|
% = \frac{1}{2} |-\mathbf{k}| = 1/2. 
% \]
% }{
% \vfill
% }

\end{document}
